Implementing Equational Constraints\\ in a Functional Language

### 2011-06-30 papers/wlp2011/paper.tex
Functional pattern section started

The condition $c$ in rule (\ref{rule}) is typically a
conjunction of \emph{equational constraints}
of the form \code{$e_1$\,=:=\,$e_2$}.
Such a constraint is satisfiable if
both sides $e_1$ and $e_2$ are evaluable to unifiable data terms.
For instance, if the symbol ``\code{++}'' denotes the usual list
concatenation operation, we can define an operation \code{last}
that computes the last element \code{e} of a non-empty list \code{xs}
as follows:
%
\label{ex:last}
\begin{curry}
  last xs | ys++[e]=:=xs = e   where ys,e free  
\end{curry}
%
As in Haskell, the rules defining most functions are
\emph{constructor-based} \cite{ODonnell85}, in (\ref{rule}) $t_1
\ldots t_n$ are made of variables and/or data constructor symbols
only.  However, Curry also allows \emph{functional pattern}
\cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might contain calls to
defined operations.  For instance, we can define \code{last} also
as:
%
\begin{curry}
  last (xs++[e]) = e 
\end{curry}
%
Here, the functional pattern \code{(xs++[e])}
states that \code{last} is reducible
to \code{e} provided that the argument can be matched against
some value of \code{(xs++[e])} where \code{xs} and \code{e} are
free variables.
By instantiating \code{xs} to arbitrary lists, the value
of \code{(xs++[e])} is any list having \code{e} as its last element.
Functional patterns are
a powerful feature to express arbitrary selections in term structures.
For instance, functional patterns supports a straightforward
processing of XML data with incompletely specified
or evolving formats \cite{Hanus11ICLP}.

@@

As known from logic programming, predicates or constraints
are important to restrict the set of intended values
in a non-deterministic computation.
Apart from user-defined predicates,
equational constraints of the form \code{$e_1$\,=:=\,$e_2$}
are the most important kind of constraints.
We have already seen in the first rule defining operation
\code{last} in Section~\ref{ex:last} a typical application
of an equational constraint.

@@

Due to the extensions described in this section,
we obtain the implementation of a unification procedure
that integrates the demand-driven evaluation of operations
occurring in the arguments.
Benchmarks evaluating the efficiency of this approach are
shown in Section~\ref{sec:Benchmarks}.

@@

>>>>

The condition $c$ in rule (\ref{rule}) is typically a
conjunction of \emph{equational constraints}
of the form \code{$e_1$\,=:=\,$e_2$}.
Such a constraint is satisfiable if
both sides $e_1$ and $e_2$ are evaluable to unifiable data terms.
For instance, if the symbol ``\code{++}'' denotes the usual list
concatenation operation, we can define an operation \code{last}
that computes the last element \code{e} of a non-empty list \code{xs}
as follows:
%
\label{ex:last}
\begin{curry}
  last xs | ys++[e]=:=xs = e   where ys,e free  
\end{curry}
%
As in Haskell, the rules defining most functions are
\emph{constructor-based} \cite{ODonnell85}, in (\ref{rule})
$t_1,\ldots,t_n$ are made of variables and/or data constructor symbols
only.  However, Curry also allows \emph{functional pattern}
\cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might contain calls to
defined operations.  For instance, we can define the last element
of a list also by
as:
%
\begin{curry}
  last' (xs++[e]) = e 
\end{curry}
%
Here, the functional pattern \code{(xs++[e])}
states that \code{last'} is reducible
to \code{e} provided that the argument can be matched against
some value of \code{(xs++[e])} where \code{xs} and \code{e} are
free variables.
By instantiating \code{xs} to arbitrary lists, the value
of \code{(xs++[e])} is any list having \code{e} as its last element.
Functional patterns are
a powerful feature to express arbitrary selections in term structures.
For instance, functional patterns supports a straightforward
processing of XML data with incompletely specified
or evolving formats \cite{Hanus11ICLP}.

@@

As known from logic programming, predicates or constraints
are important to restrict the set of intended values
in a non-deterministic computation.
Apart from user-defined predicates,
equational constraints of the form \code{$e_1$\,=:=\,$e_2$}
are the most important kind of constraints.
We have already seen in the operation
\code{last} in Section~\ref{ex:last} a typical application
of an equational constraint.

@@

Due to the extensions described in this section,
we obtain an implementation of a unification procedure
that integrates the demand-driven evaluation of operations
occurring in the arguments.
Benchmarks evaluating the efficiency of this approach are
shown in Section~\ref{sec:Benchmarks}.

@@

A well-known disadvantage of equational constraints
is the fact that \ccode{=:=} is interpreted as strict equality.
Thus, if one uses equational constraints to express
requirements on arguments, the resulting operations might be too strict.
For instance, the equational constraint in the condition
defining \code{last} (see Section~\ref{ex:last})
requires that \code{ys++[e]} as well as \code{xs}
must be reducible to identical values so that the input list \code{xs}
is completely evaluated.
Hence, if \code{failed} denotes an operation whose evaluation fails,
the evaluation of \code{last [failed,2]} has no result.
On the other hand, the evaluation of \code{last' [failed,2]} yields
the value \code{2}, i.e., the definition of \code{last'} is less strict.
This is due to the fact that a functional pattern like
\code{(xs++[e])} abbreviates all values to which it can be evaluated
(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]} etc.
Conceptually, the rule defining \code{last'}
abbreviates the following (infinite) set of rules:
\begin{curry}
  last' [e] = e
  last' [x1,e] = e
  last' [x1,x2,e] = e
  $\ldots$
\end{curry}
As a further example,
consider an operation that returns the first duplicated element in
a list.
Using equational constraint, we can define it as follows:
\begin{curry}
  fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
            = e    where ys,zs,e free
\end{curry}
The first equational constraint is used to split the input list \code{xs}
into three sublists.
The last constraint ensures that the first sublist
\code{ys} does not contain duplicated elements
(the library operation removes all duplicates from a list)
and the second constraint ensures that the first element after \code{ys}
occurs in \code{ys}.
Although this implementation is concise, it cannot be applied
to infinite lists due to the strict interpretation of \ccode{=:=}.
This is not the case if we define this operation by a functional pattern:
\begin{curry}
  fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
                        = e
\end{curry}
%
Functional patterns are quite useful in situations
where one wants to select parts of some structure
at arbitrary positions. For instance,
functional patterns have been applied to support
the high-level programming of querying or transforming
XML data \cite{Hanus11ICLP}.

Obviously, one cannot implement functional patterns by
a transformation into an infinite set of rules.
Fortunately, functional patterns can be implemented by
a specific \emph{lazy unification} procedure \ccode{=:<=}
\cite{AntoyHanus05LOPSTR}.
For instance, the definition of \code{last'} is transformed into
\begin{curry}
  last' ys | (xs++[e]) =:<= ys
           = e   where xs,e free
\end{curry}
The behavior of \ccode{=:<=} is similar to \ccode{=:=}
except for the case that a variable in the left argument
should be bound to some expression: instead of evaluating
the expression to some value and bind the variable to it,
the variable is bound to the \emph{unevaluated} expression
(see \cite{AntoyHanus05LOPSTR} for more details).
Due to this slight change, failures or infinite structures
in actual arguments do not cause problems in the matching
of functional patterns.

This discussion shows the requirements on an implementation
of functional patterns. The general structure
is quite similar to equational constraints
with the exception that variables could be also bound
to unevaluated expressions. If such variables are later
accessed, they are evaluated.
This can be achieved by adding a further alternative
to the type of decisions:
\begin{haskell}
data Decision = $\ldots$ | LazyBind [Constraint]
\end{haskell}
%
The implementation of the lazy unification \ccode{=:<=}
is almost identical to the strict unification \ccode{=:=}
